### 2018-02-02 thesis/contents/reviews/pain.tex
Updates from ShareLaTeX

Trigeminal Neuralgia is a chronic neuropathic facial pain syndrome. The quest for the understanding of the pathophysiology of this syndrome requires a review of the history and theories of pain. Pain is an unique and important sensory dimension that is central to the evolution and survival of animal species. In humans, the sensation of pain is typically described as an unpleasant quality. The experience of pain is complex, consisting of interaction between nociceptive, emotional and cognitive processes. The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\footnote{http://www.iasp-pain.org/Taxonomy\#Pain}.

@@

 \begin{figure}[ht]
 \includegraphics[width=\textwidth]{pain-pathways.jpg}
 \centering
 \caption{Afferent pain pathways} 
 \label{fig:pain-pathways}
 \end{figure}

@@

\paragraph{Sensory-Discrimination}
Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the primary sensory (S1) and secondary sensory (S2) cortices. The S1 cortex is responsible for the discrimination of fine-touch sensations. S1 receives inputs directly from VP thalamus. Nonetheless, S1 has not been consistently reported in acute pain studies, possibly due to the difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within S1 \cite{bushnell1999pain}. S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies. The posterior insula also receives direct nociceptive inputs from the VMPo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}. Therefore it is believe to plays an important role in pain sensory-discrimination.

\paragraph{Affective-Motivational}
The cortical regions that are involved in affective-motivational pain procecssing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions. The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the insula receives input from the VMpo thalamus. The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain. These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}.

>>>>

Trigeminal Neuralgia is a chronic neuropathic facial pain syndrome. The quest for the understanding of the pathophysiology of this syndrome requires a review of the history and theories of pain. Pain is an unique and important sensory dimension that is central to the evolution and survival of animal species. In humans, the sensation of pain is typically described as an unpleasant quality. The experience of pain is complex, consisting of interaction between nociceptive, emotional and cognitive processes. The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\cite{Merskey2012}.

@@

 \begin{figure}[ht]
 \includegraphics[width=\textwidth]{pain-pathways.jpg}
 \centering
 \caption{Afferent pain pathways \protect\cite{kandel2000principles}}
 \label{fig:pain-pathways}
 \end{figure}

@@

\paragraph{Sensory-Discrimination}
Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the primary sensory (S1) and secondary sensory (S2) cortices. The S1 cortex is responsible for the discrimination of fine-touch sensations. S1 receives inputs directly from VP thalamus. Nonetheless, S1 has not been consistently reported in acute pain studies, possibly due to the difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within S1 \cite{bushnell1999pain}. S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies \cite{Wager2013,Brodersen2012}. The posterior insula also receives direct nociceptive inputs from the VMPo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}. Therefore it is believe to plays an important role in pain sensory-discrimination.

\paragraph{Affective-Motivational}
The cortical regions that are involved in affective-motivational pain processing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions. The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the insula receives input from the VMpo thalamus. The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain. These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}. The unpleasantness of pain has been thought to be a motivational force \cite{Auvray2010}, that biases actions towards avoidance. There is evidence that the motor cortex, which is involved in action planning, and the ACC are activated during pain \cite{Erpelding2013}. The cortical excitability of the motor cortex is also shown to be altered in chronic pain \cite{Parker2016}. The motivational system is also involved in pain-reward and involves the ACC, lateral orbitofrontal cortex (OFC), and the prefrontal cortex (PFC, both ventromedial and anterior aspects) \cite{Navratilova2015}. The nucleus accumbens (NAc) BOLD activity correlated with anticipation of pain relief \cite{Baliki2010}. The reward/motivational circuitry is thought to be disrupted in chronic pain, and such disruption is related to the emotional and cognitive deficits in chronic pain sufferers.

### 2018-02-02 thesis/contents/reviews/trigeminal.tex
Updates from ShareLaTeX

 \begin{figure}[ht]
 \includegraphics[width=120mm]{trig-branches.png}
 \centering
 \caption{Anatomy of the trigeminal nerve branches.} 
 \label{fig:trig-branches} 
 \end{figure}

@@

Classic TN is believed to be caused by neurovascular compression (NVC) of the CN V at its root entry zone (REZ) by the surrounding vasculature, where the nerve interfaces with the pons. An aberrant loop of artery and less commonly a vein is seen in 80-90\% of the cases of TN compressing the nerve \cite{Love2001,McLaughlin1999}. This can also be observed in conventional MR imaging \cite{Borges2010}. The superior cerebellar artery is responsible for most of the NVC (60-90\%), followed by the anterior inferior cerebellar artery (AICA) and basilar arteries \cite{Lutz2011}. There is a slight lateral prevalence, with right-sided TN occurring in  in 60\% of the cases, and in rare cases may be related to asymmetries of the foramen rotundum and foramen ovale \cite{Toda2009}.  Other types of compression may also cause TN; these can be in the forms of vestibular schwannomas, meningiomas, epidermoid cysts, and other cerebellopontine angle tumours, arteriovenous malformations and saccular aneurysms . However NVC cannot be assumed to be directly causal of TN, since observation of vascular compression does not necessarily result in the clinical expression of pain \cite{Desouza2013,Hodaie2013}. In approximately 10-15\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006,Maarbjerg2015}.

@@

Neuroimaging of the TN has been challenging, due to the small diameter of the CN V and since the contrast of the CNS trigeminal pathways is difficult to distinguish from the surrounding on conventional MRI \cite{Borges2010}. Manual segmentation of T1 and T2 MRI requires great effort, with mixed results \cite{Miller2008a}. Diffusion MRI provided fiber orientation, and early attempts to segment the trigeminal system showed promise \cite{Upadhyay2008,Habas2007f}.

>>>>

 \begin{figure}[ht]
 \includegraphics[width=120mm]{trig-branches.png}
 \centering
 \caption{Anatomy of the trigeminal nerve branches. \protect\cite{gray1918anatomy}} 
 \label{fig:trig-branches} 
 \end{figure}

@@

Classic TN is believed to be caused by neurovascular compression (NVC) of the CN V at its root entry zone (REZ) by the surrounding vasculature, where the nerve interfaces with the pons. An aberrant loop of artery and less commonly a vein is seen in 80-90\% of the cases of TN compressing the nerve \cite{Love2001,McLaughlin1999}. This can also be observed in conventional MR imaging \cite{Borges2010}. The superior cerebellar artery is responsible for most of the NVC (60-90\%), followed by the anterior inferior cerebellar artery (AICA) and basilar arteries \cite{Lutz2011}. There is a slight lateral prevalence, with right-sided TN occurring in  in 60\% of the cases, and in rare cases may be related to asymmetries of the foramen rotundum and foramen ovale \cite{Toda2009}.  Other types of compression may also cause TN; these can be in the forms of vestibular schwannomas, meningiomas, epidermoid cysts, and other cerebellopontine angle tumours, arteriovenous malformations and saccular aneurysms \cite{Haller2016,Love2001}. However NVC cannot be assumed to be directly causal of TN, since observation of vascular compression does not necessarily result in the clinical expression of pain \cite{Desouza2013,Hodaie2013}. In approximately 10-15\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006,Maarbjerg2015}.

@@

Neuroimaging of the TN has been challenging, due to the small diameter of the CN V and since the contrast of the CNS trigeminal pathways is difficult to distinguish from the surrounding on conventional MRI \cite{Borges2010}. In conventional MRI, the brainstem nerve segment can only be deduced from its surrounding structures \cite{Casselman2008}. Sequences that can visualize myelinated structures are chosen, these include T2-weighted MRI  sequences such as FIESTA and CISS. Manual segmentation of T1 and T2 MRI requires great effort, with mixed results \cite{Miller2008a}. Diffusion MRI provided fiber orientation, and early attempts to segment the trigeminal system showed promise \cite{Upadhyay2008,Habas2007f}. Additionally, DTI showed strong ability to visualize spatial relationships in white matter anatomy, for example in delineating fornix sub-regions \cite{Chen2015c}, and in visualizing the distortion of CN VII/VIII by vestibular schwannomas \cite{Chen2011b}.

### 2018-02-02 thesis/contents/tn_sagit.tex
Updates from ShareLaTeX

\chapter{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
\chaptermark{Study IV}

\section{Abstract}
\textbf{Introduction:} Trigeminal Neuralgia (TN) is a debilitating facial neuropathic pain syndrome involving the trigeminal sensory pathway. We demonstrate the use of a fully-automated group white-matter tractography, and Gaussian Process (GP) classification to pinpoint key white-matter diffusivity changes that maximally differentiate patients and healthy controls.

\textbf{Methods:} We performed group merged tractography using SAGIT on 36 sex-matched TN patients (right-sided pain) and 36 controls. The trigeminal nerve (CN V), pontine decussation, and thalamocortical S1 pathways were reconstructed. GP classifiers were trained by scrolling a moving window over CN V and S1 tractography centroids. FA, GFA, RD, AD, and MD metrics were evaluated. Both TN vs. Control and Affect vs. Unaffected were analyzed. Classifiers that performed at greater-or-equal-to 75\% accuracy were accepted.

@@

\textbf{Conclusions:} We have demonstrated the first instance of group-wise merged tractography analysis of TN vs. control,  as well as the first white-matter machine learning classification with GP. The method reproduced previous findings with finer-grained regional specificity. The study also revealed specific diffusivity changes in deeper white matter.

\section{Introduction}
\subsubsection{Trigeminal Neuralgia}
Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more dermatomes of the three trigeminal nerve (The fifth cranial nerve: CN V) branches. TN manifests classically in idiopathic fashion, or secondary to a range of neuropathology, including meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}. The pathophysiology of idiopathic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidence of vascular compression \cite{Lee2014}. We previously compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is disrupted in MS-TN. In all previous TN studies, placements of region-of-interests in the CN V were pre-determined and manually placed. In this study we further expand the study to A) analyze the CN V differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interest; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.

The trigeminal sensory system involves both the discriminative touch and pain pathways. The discriminative pathway synapsis onto the primary trigeminal sensory nucleus, and then project to the contralateral VC thalamus, and then to the S1 cortex as we have delineated. The pain pathway synapsis caudally onto the trigeminal spinal nucleus, and decussate at the level of the medulla and spinal cord. The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus. The pain pathway also projects to the medial nucleus of the thalamus, where it then project medially to the cingulate cortex.

\subsubsection{Data Analysis from Streamlines}
Neuroimaging tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}. To obtain diffusivity metrics from tractography involves converting the streamline bundles into a volumetric spatial mask, and then measure the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}. However this approach discards the orientation information in the streamlines. The naive form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-through. This becomes an issue when considering that an average 3-tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm. In white matter structures such as CN V, and other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.

Consider that a tractography streamline propagation algorithm involves all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}. It is reasonable to consider these algorithms to be an interpolated subsampling scheme of the DWI volume that takes the diffusion-based orientation into account. Therefore, it is desirable to directly make use of the tractography streamlines for data measurement.

Along-the-tract analysis has been investigated by various authors \cite{Colby2012,ODonnell2009,Wang2015,Yeatman2012}. In these approaches, the bundles are selectively filtered from global tractography \cite{Wang2015,Yeatman2012}, and studies were limited to major white matter bundles such as the corpus callosum, corticospinal tract, longitudinal fasciculus, arcuate fasciculus, and uncinate. The measures were pooled onto tract centroid in each individual subjects, and then aggregated at the group level. The global tractography approach is unintuitive for anatomy-specific tasks such as neurosurgical planning, where the anatomy of interest is clear. A more focused ROI based delineation approach is therefore preferred in these circumstances.

It is important to note that streamline clustering and centroid identification can be considered as independent tasks. Some clustering methods do not provide a best-representative streamline centroid. Centroid identification involved resampling the streamlines into an identical number of points, and then identifying point correspondences via different distance measures. For these approaches, it is vital that the streamlines have similar length, and an assumption that the start and end of the streamlines across subjects are consistent. However, pathological brain tissue may negatively affect tractography delineation. Therefore the geometry assumptions of such approaches may not be satisfied when extending them to study patient populations. Running time and memory constraints also need to be taken into consideration, when applying such algorithms at a group level. One method for tract clustering and centroid identification that is shown to have linear running time in respect to streamline count is QuickBundles \cite{Garyfallidis2012}. We have adopted QuickBundles to obtain S1 streamline centroids.

The principal advantage of a streamline-based approach is that we can retain the pathway orientation from its geometry. White matter metrics in 3D space can then be projected to a lower dimensional centroid for detailed analysis. We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework (Chen et al., 2016) to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning. We then merge the tracts into a template space using non-linear deformations, and then perform tractography centroid identification and statistics at the merged group level to ensure anatomical consistency.

\subsubsection{Gaussian Process}
The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} have sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}. However, one disadvantage of DNN is in its inability to quantify uncertainty and provide interpretable models. Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process (GP) \cite{gal2016dropout}. In fact, it has been found that DNN with an infinite number of hidden units approximates GP \cite{neal1996priors}.

Gaussian Process is defined \cite{rasmussen2006gaussian}) as a collection of random variables, any finite number of which have a joint Gaussian distribution. 
It is a distribution that is defined by a mean function $m(x)$ and covariance function $k(x,x') $ where
\begin{equation}
	\begin{split}
		m(x) &= \mathbb{E}[f(x)], \\
		k(x,x') &= \mathbb{E}[(f(x)-m(x)(f(x')-m(x'))],
	\end{split}
\end{equation}
And the Gaussain processed is defined as
\begin{equation}
f(x) \sim GP(m(x), k(x, x')) 
\end{equation}

The covariance function is also termed called kernel function. The radial-basis function kernel or RBF kernel is widely used in machine learning tasks. GP has previously been explored in neuroimaging for tractography clustering \cite{Wassermann2010}. It is particularly well suited for continuous data streams, such as tractography streamlines.

\section{Methods}
\subsubsection{Acquisitions}
The study recruited 36 idiopathic TN patients (28 females, 8 males; age 60.1$\pm$16 years; all right side pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).  All TN patients were scanned at 1 day pre-GK and 6 months post-GK. MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions. T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256x256. DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.

\subsubsection{Preprocessing}
DWI images were motion/eddy-current corrected, with appropriate correction applied to the b-matrix \cite{Leemans2009}. Region of interests (ROI) for tractography seeding were manually created on a normalized template brain from 42 healthy brains. The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}.

We deployed selective automatic group integrated tractography (SAGIT) framework to automate preprocessing, registration, and tractography. The processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations. The T1 images for all the subjects were registered to the template using symmetric diffeomorphic registration (SDR) \cite{Avants2008b}. Each individual DWI was registered to their T1 image with SDR, using mean DWI image for DWI-T1 co-registration. Freesurfer segmentation maps were first affine transformed from the Freesurfer space to the T1 space using FMRIB's Linear Image Registration Tool (FLIRT) \cite{Jenkinson2001,Jenkinson2002}, and then transformed using SDR transforms to the individual DWI space for selective ROI generation with SAGIT.

\subsubsection{Tractography}
ROIs were defined on the template image to bilaterally delineate: 1) CN V (ROI: cistern nerve root; excludes: cerebellum grey matter, pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs: trigeminal nucleus (TGN); includes: contralateral VC thalamus, pontine decussation; excludes: ipsilateral thalamus, cerebellar grey matter, pons, contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs: VC thalamus; includes: ipsilateral S1 white-matter boundary; excludes: brainstem).

@@

Tractography was first performed in the native DWI space for each individual.  FA, AD, RD, MD, and GFA metrics were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation. Additionally, subject ID and group ID scalar values were also inserted into each tractography files for identification.

\subsubsection{Merged Tractography}
All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis. A software library called “Fascicles” is created by the author to facilitate this process. The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and then all points were exported into a comma-delimited file. The points were then transformed by applying the subject DWI to T1, and T1 to template transforms using ANTs’  applyTransformToPoints command \cite{Avants2009}. The transformed points were then imported back into the SQLite database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group ID associations preserved. Finally, all template-space tractography models across all the subjects are imported into a new SQLite database, and then exported as a single file for analysis.

@@

Tractography centroids, which were the average curve best fit the tract bundle, was defined on the merged tracts in template space. CN V centroids were manually defined. The CN V centroids were defined by a list of 3 control points situated in the template world space. A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}, which resulted in a curve of length 38.91 mm. The spline is then resampled into 30 points that become the centroid definition with segment length of 1.3 mm. Nearest-neighbor point assignment was used to divide the merged tract into 30 cross sections along the centroids. Diffusion metrics from the points were aggregated for each of the cross-sections.  S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30, with centroid length of 62.1 mm and segment length of 2.07 mm.

@@

The diffusivity measures of CN V reproduced our previous findings \cite{Chen2016a}, which was that cistern, REZ differentiated TN from controls. The consistent regional overlaps in the cistern/REZ region of the right (affected) CN V in MD, RD, AD shows the ability for the GP classifiers to auto-converge to focal white-matter abnormality. The left versus right GP classifier placed left-right differences at the cistern segment, in agreement with the pathophysiology of TN. The regional findings on left versus controls suggest that there is bilateral differences in CN V myeline in unilateral TN, despite the lack of symptoms on the unaffected side. The left regions also situate into the brainstem segments. It is possible that the left CN V is affected by global descending pain modulation, and warrant further investigation.

>>>>

\chapter{Study IV: Gaussian Process Classification of Trigeminal Neuralgia using Merged Tractography}
\chaptermark{Study IV}
\label{section:study4}

\section{Abstract}
\textbf{Introduction:} Trigeminal Neuralgia (TN) is a debilitating facial neuropathic pain syndrome involving the trigeminal sensory pathway. We demonstrate the use of a fully-automated group white-matter tractography and Gaussian Process (GP) classification to pinpoint key white-matter diffusivity changes that maximally differentiate patients and healthy controls.

\textbf{Methods:} We used SAGIT based group merged tractography for 36 sex-matched TN patients (right-sided pain) and 36 controls, examining the following white matter pathways: trigeminal nerve (CN V), pontine decussation, and thalamocortical fibers. GP classifiers were trained by scrolling a moving window over CN V and S1 tractography centroids. FA, GFA, RD, AD, and MD metrics were evaluated for both groups, analysing TN vs. Control and affected vs. unaffected groups. Classifiers that performed at greater-or-equal-to 75\% accuracy were accepted.

@@

\textbf{Conclusions:} We  demonstrate the first instance of group-wise merged tractography analysis of TN vs. control and TN white-matter machine learning classification using GP. The method reproduced previous findings with finer-grained regional specificity. The study also revealed specific diffusivity changes in deeper white matter.

\section{Introduction}
\subsubsection{Trigeminal Neuralgia}
Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more dermatomes of the three trigeminal nerve (The fifth cranial nerve: CN V) branches. TN manifests most commonly as classic (also called idiopathic TN), or secondary to a range of neuropathology, including tumors such as meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}. The pathophysiology of idiopathic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidence of vascular compression \cite{Lee2014}. We previously compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is disrupted in MS-TN. In all previous TN studies, placements of region-of-interests in the CN V were pre-determined and manually placed. In this study we further expand the study to A) analyze the CN V differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interest; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.

The trigeminal sensory system involves both the discriminative touch and pain pathways. Anatomicall, the discriminative pathway synapses onto the primary trigeminal sensory nucleus, and then projects to the contralateral VC thalamus and then to the S1 cortex. The pain pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord. The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus. The pain pathway also projects to the medial nucleus of the thalamus, where it then project medially to the cingulate cortex.

\subsubsection{Data Analysis from Streamlines}
White matter tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}. Obtaining diffusivity metrics from tractography involves converting streamline bundles into a volumetric spatial mask, followed by measurement of the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}. However this approach discards the orientation information in the streamlines. The naive form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-through. This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm. In white matter structures such as CN V as well as other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.

In this light, it must be considered that a tractography streamline propagation algorithm involves all or parts of the following stages: 1) subsampling of an initial region for seeding \cite{Basser2002,Cote2012}; 2) identifying a propagation direction, based on previously traversed paths \cite{Malcolm2010,Qazi2009,Tournier2010}; 3) smoothing of the traversed paths for output \cite{Tuch2000d}. It is reasonable to consider these algorithms to be an interpolated subsampling scheme of the DWI volume that takes the diffusion-based orientation into account. Therefore, it would be desirable to directly make use of the tractography streamlines for data measurement.

Along-the-tract analysis has been investigated by various authors \cite{Colby2012,ODonnell2009,Wang2015,Yeatman2012}. In these approaches, the bundles were selectively filtered from global tractography \cite{Wang2015,Yeatman2012}, and limited to major white matter bundles such as the corpus callosum, corticospinal tract, longitudinal fasciculus, arcuate fasciculus, and uncinate. The measures were pooled onto tract centroid in each individual subjects, and then aggregated at the group level. The global tractography approach is unintuitive for anatomy-specific tasks such as neurosurgical planning, where the anatomy of interest is clear. A more focused ROI based delineation approach is therefore preferred in these circumstances.

It is important to note that streamline clustering and centroid identification can be considered as independent tasks. Some clustering methods do not provide a best-representative streamline centroid. Centroid identification involves resampling the streamlines into an identical number of points, and then identifying point correspondences via different distance measures. For these approaches, streamlines require similar length, and an assumption that the start and end of the streamlines across subjects is consistent. However, pathological brain tissue may negatively affect tractography delineation. Therefore the geometry assumptions of such approaches may not be satisfied when studying patient populations. Running time and memory constraints also need to be taken into consideration when applying such algorithms at a group level. We have adopted QuickBundles since it is a practical method of tract clustering and centroid identification with linear running time to obtain S1 streamline centroids. \cite{Garyfallidis2012}

The principal advantage of a streamline-based approach is that we can retain the pathway orientation from its geometry. White matter metrics in 3D space can then be projected to a lower dimensional centroid for detailed analysis. We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework (Chen et al., 2016) to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning. We then merge the tracts into a template space using non-linear deformations, and  perform tractography centroid identification and statistics at the merged group level to ensure anatomical consistency.

\subsubsection{Gaussian Process}
The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} have sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}. However, one disadvantage of these networks is in its inability to quantify uncertainty and provide interpretable models. Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process (GP) \cite{gal2016dropout}. In fact, it has been found that DNN with an infinite number of hidden units approximates GP \cite{neal1996priors}.

Gaussian Process is defined \cite{rasmussen2006gaussian} as a collection of random variables, any finite number of which have a joint Gaussian distribution. 
It is a distribution that is defined by a mean function $m(x)$ and covariance function $k(x,x') $ where
\begin{equation}
	\begin{split}
		m(x) &= \mathbb{E}[f(x)], \\
		k(x,x') &= \mathbb{E}[(f(x)-m(x)(f(x')-m(x'))],
	\end{split}
\end{equation}
And the Gaussain processed is defined as
\begin{equation}
f(x) \sim GP(m(x), k(x, x')) 
\end{equation}

The covariance function is termed called kernel function. The radial-basis function kernel (RBF kernel) is widely used in machine learning tasks. GP has previously been explored in neuroimaging for tractography clustering \cite{Wassermann2010} and it is particularly well suited for continuous data streams, such as tractography streamlines.

\section{Methods}
\subsubsection{Acquisitions}
We studied 36 patients with classic TN (28 females, 8 males; age 60.1$\pm$16 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).  All TN patients were scanned at 1 day pre-GK and 6 months post-GK. MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions. T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256x256. DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.

\subsubsection{Preprocessing}
DWI images were motion/eddy-current corrected, with appropriate correction applied to the b-matrix \cite{Leemans2009}. Region of interests (ROI) for tractography seeding were manually created on a normalized template brain from 42 healthy brains. The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}. We deployed selective automatic group integrated tractography (SAGIT) framework to automate preprocessing, registration, and tract generation. The processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations. The T1 images for all the subjects were registered to the template using symmetric diffeomorphic registration (SDR) \cite{Avants2008b}. Each individual DWI was registered to their T1 image with SDR, using mean DWI image for DWI-T1 co-registration. Freesurfer segmentation maps were first affine transformed from the Freesurfer space to the T1 space using FMRIB's Linear Image Registration Tool (FLIRT) \cite{Jenkinson2001,Jenkinson2002}, and then transformed using SDR transforms to the individual DWI space for selective ROI generation with SAGIT.

\subsubsection{Tractography}
ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (ROI: cistern nerve root; excludes: cerebellum grey matter, pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs: trigeminal nucleus (TGN); includes: contralateral VC thalamus, pontine decussation; excludes: ipsilateral thalamus, cerebellar grey matter, pons, contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs: VC thalamus; includes: ipsilateral S1 white-matter boundary; excludes: brainstem).

@@

Tractography was first performed in the native DWI space for each subject. Diffusivity metrics of FA, AD, RD, MD, and GFA  were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation. Additionally, subject ID and group ID scalar values were also inserted into each tractography files for identification.

\subsubsection{Merged Tractography}
All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis. A software library called “Fascicles” is created by the author to facilitate this process. The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported into a comma-delimited file. The points were then transformed by applying the subject DWI to T1, and T1 to template transforms using ANTs’  applyTransformToPoints command \cite{Avants2009}. The transformed points were then imported back into the SQLite database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group ID associations preserved. Finally, all template-space tractography models across all the subjects are imported into a new SQLite database, and then exported as a single file for analysis.

@@

Tractography centroids, which were the average curve best fit the tract bundle, was defined on the merged tracts in template space. CN V centroids were manually defined, using 3 control points situated in the template world space. A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}, which resulted in a curve of length 38.91 mm. The spline is then resampled into 30 points that become the centroid definition with segment length of 1.3 mm. Nearest-neighbor point assignment was used to divide the merged tract into 30 cross sections along the centroids. Diffusion metrics from the points were aggregated for each of the cross-sections.  S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30, with centroid length of 62.1 mm and segment length of 2.07 mm.

@@

The diffusivity measures of CN V reproduced our previous findings \cite{Chen2016a}, which was that cistern, REZ differentiated TN from controls. The consistent regional overlaps in the cistern/REZ region of the right (affected) CN V in MD, RD, AD shows the ability for the GP classifiers to auto-converge to focal white-matter abnormality. The left versus right GP classifier placed left-right differences at the cistern segment, in agreement with the pathophysiology of TN. The regional findings on left versus controls suggest that there is bilateral differences in CN V myelin in unilateral TN, despite the lack of symptoms on the unaffected side. The left regions also situate into the brainstem segments. It is possible that the left CN V is affected by global descending pain modulation, and warrant further investigation.
